---
title: "Passing expressions and data from R to C++ at compile-time in Rmarkdown"
author: "Edger Sterjo"
date: 2020-06-21
categories: ["C++", "R"]
tags: ["metaprogramming", "rmarkdown", "c++", "rcpp", "r"]
---


## Introduction

In this post we give a simple illustrative example of how data generated by R code can be used by compiled languages such as C++ at compile time, instead of run-time, inside Rmarkdown.

This is an example of inter-language code generation. Metaprogramming/code generation is an extremely powerful technique but it's also one that is very easy to overdo. This is just a fun example to learn from. Thorough testing is very important for any production code.


## Using other languages in Rmarkdown

Out of the box Rmarkdown can work with the following languages assuming a proper back-end is available:

```{r}
names(knitr::knit_engines$get())
```
Although we can use R's native foreign function interface to call compiled code, for C++ a higher level alternative is to use [Rcpp](https://cran.r-project.org/web/packages/Rcpp/index.html). In Rmarkdown we can [compile C++ code chunks using Rcpp and export the compiled functions to be available for use in R](https://bookdown.org/yihui/rmarkdown/language-engines.html#rcpp).

As a common example, we can compile the following code
```{Rcpp my_chunk, cache=TRUE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector timesTwo(NumericVector x) 
{
	return x * 2;
}
```
and use the exported function in R
```{r}
timesTwo(1:10)
```


## Registering a user-defined language engine in Knitr

We can create [user-defined engines](https://bookdown.org/yihui/rmarkdown-cookbook/custom-engine.html) to control exactly how the code chunk is sourced, or even modify existing engines. To get an idea we can look at the default Rcpp engine used by knitr:

```{r}
knitr::knit_engines$get()$Rcpp
```

Using the default engine above as a template we can define a new knitr engine for compiling C++. One that can read and make use of more dynamic R data in C++ before compilation (or even dynamically create `Makevars` files to control compilation flags). First let's include the knitr package:


```{r}
library(knitr)
```

Next let's take a crack at defining a new engine to compile C++ code. In this example we will modify the current Rcpp engine to take in an `extra` field (but otherwise behave the same).
```{r}
knit_engines$set(RcppFoo = function(options) {
	
	extra = options$extra
	
	sourceCpp = getFromNamespace("sourceCpp", "Rcpp")
	
	## Code is read as a list of strings, one list element per line
	## Here we append extra code that may be defined in R to the 
	## code written in the chunk
	code = c(extra, options$code)
    code = paste(code, collapse = '\n')
    opts = options$engine.opts
    
    if (!is.environment(opts$env)) 
        opts$env = knit_global()

    if (options$eval) {    
		message("Building shared library for Rcpp code chunk...")
		do.call(sourceCpp, c(list(code = code), opts))
    }
    options$engine = "cpp"
    engine_output(options, 
    			  options$code, 
    			  paste("Added the lines:\n", 
    			  	  paste(extra, collapse = '\n'), 
    			  	  sep = '\n'))
})
```

Next we test by creating some data in R and using that as a compile time constant in C++. Here we pass values of pi and e as static consts to C++.

```{r}
constants = list(
	paste('static const double Pi =', pi, ';'),
	paste('static const double Euler =', exp(1),';')
)
```
This already highlights a danger as we have not considered exactly how R might convert these double precision floating point numbers to strings. Regardless, we proceed. To use the new engine we run the engine as `{RcppFoo test_chunk, extra = constants}`

```{RcppFoo test_chunk, extra=constants}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector timesFoo(NumericVector x) 
{
	return x * Pi + Euler;
}
```

```{r}
x = timesFoo(1:10)
print(x)
```
We get **almost** the same result as in R
```{r}
y = pi*(1:10)+exp(1)
print(y)
```
But metaprogramming can be dangerous. In this case some loss of precision occurred with the doubles when converting to strings:
```{r}
print(x - y)
```
Anyway this was just a small example. There are many many directions one can choose to take with metaprogramming. Even creating new preprocessing directives such as unrolling loops, defining constexprs, etc.