---
title: "Expectation Maximization, Part 2: Illustrating Bias-Variance Trade Off of Regularization with Probit Regression"
author: "Edger Sterjo"
date: 2020-08-01
categories: ["R","C++"]
tags: ["r","rcpp","eigen","c++","expectation maximization", "em algorithm","em","regularization", "machine learning", "statistical learning", "bias variance tradeoff"]
---

## Introduction

In the [first post](https://edsterjo.netlify.app/post/003_em1/main/) in this series we discussed Expectation Maximization (EM) type algorithms. In the post [prior to this one](https://edsterjo.netlify.app/post/004_regularization/main/) we discussed regularization and showed how it leads to a bias-variance trade off in OLS models. Here we implement fitting for $L^2$-regularized [probit regression](https://en.wikipedia.org/wiki/Probit_model) using EM. To make it more interesting we will code everything from scratch using the [Eigen](http://eigen.tuxfamily.org/index.php?title=Main_Page) linear algebra library, via [RcppEigen](https://cran.r-project.org/web/packages/RcppEigen/index.html).


## Probit Regression as a censored OLS model

In our first post on EM algorithms we emphasized that EM is particularly useful for models that have censored data. Suppose we have the following censored model. Suppose $Y^*$ is a real valued random variable and $\vec{X}$ is a random vector with values in $\mathbb{R}^p$. Suppose that we have the conditional relationship

$$
Y^* \ \ | \ \ \vec{X} \sim \mathcal{N}(\ \langle \vec{X}  , \beta\rangle \ ,\ 1)
$$

where $\mathcal{N}(\mu, \sigma^2)$ denotes the univariate normal distribution with mean $\mu$ and variance $\sigma^2$. Here (and everywhere else) the symbol $\langle v,w\rangle$ represents the Euclidean inner product (aka, [dot product](https://en.wikipedia.org/wiki/Dot_product)) of two vectors $v$ and $w$.

In this case we may write $Y^* = \langle \vec{X}  , \beta\rangle -  \epsilon$, where $\epsilon \sim \mathcal{N}(0,1)$ is standard normal, which can be taken independent of $\vec{X}$ or this distribution can be simply assumed conditional on $\vec{X}$.

Instead of observing $Y^*$ in the data however, we observe the censored variable
$$
Y := 
\begin{cases}
1, & \text{if} \ \ Y^* > 0 \\
0, & \text{otherwise}
\end{cases}
$$
Hence we have that $Y \ |\vec{X} \sim \text{Bernoulli}(p)$ where
$$
p = P(Y = 1 | \vec{X}) = P(Y^* > 0 | \vec{X}) = P\bigg(Y^* - \langle \vec{X},\beta\rangle > - \langle \vec{X},\beta\rangle\ \ \bigg|\ \ \vec{X}\bigg) = P(\epsilon < \langle \vec{X},\beta\rangle\ | \vec{X})
$$
Since $\epsilon \sim \mathcal{N}(0,1)$ this last probability is equal to $\Phi(\langle \vec{X},\beta\rangle)$ where $\Phi$ denotes the standard normal CDF. This derives the probit model.

Before we proceed, notice 2 points which we won't dwell on:

* If the variance of $Y^*$ had been $\sigma^2 \ne 1$ then the value of $\sigma$ would not be able to be estimated from data without knowing $\beta$ since $P(Y^* > 0 |\vec{X}) = P(Y^*/\sigma > 0 | \vec{X})$.
* If the distributional relationship between $Y$ and $\vec{X}$ had been such that the error term $\epsilon$ where a logistic random variable, instead of a normal one, then the censored problem would have become logistic regression.

For the regression itself we assume that we have a data set $\{ (y_i, \vec{x}_i)\}_{i = 1}^N$ consisting of samples generated independently of one another from a fixed multivariate distribution for $(Y, \vec{X})$ (i.e. we assume our data was sampled IID).




## Fitting Probit Regression via EM

Since probit regression arises from a censored normal OLS model, and since OLS is relatively easy to fit, probit regression is an excellent candidate for applying Expectation Maximization for fitting. A small difference will be that all of the probability densities involved will be conditional on the observed covariates $\{\vec{x}_i\}_{i = 1}^N$ since regression is a conditional relationship.

Let's recall the standard EM algorithm for the case of the regression problem above:

* **Step 1:** Let $m = 0$. Make an initial estimate $\beta_m$ for $\beta$.

* **Step 2:** Given the observed data $\{(y_i, \vec{x}_i)\}_{i = 1}^N$ and pretending for the moment that our current guess $\beta_m$ is correct, construct the conditional probability distribution $p(\{y^*_i\}\ \ |\ \ \{(y_i, \vec{x}_i)\},\ \beta_m)$ of the hidden data $\{Y^*_i\}$ given all known information.

* **Step 3:** Using the distribution $p(\{y^*_i\}\ \ |\ \ \{(y_i, \vec{x}_i)\},\ \beta_m)$ construct the following estimator/approximation of the desired log-likelihood $\log(p(\{Y^*_i\} \ | \ \beta, \{\vec{x}_i\}))$ for arbitrary values of $\beta$:

$$
Q(\beta | \beta_m) \ := \ \text{E}_{\{Y^*_i\}\ \ |\ \ \{(Y_i=y_i,\ \vec{X}_i=\vec{x}_i)\},\ \beta_m}  \big[ \log(p(\{Y^*_i\} \ | \ \beta, \{\vec{x}_i\})) \big] 
$$
$$
= \int_{\mathcal{Y^*}} \log(p(\{y^*_i\} \ | \ \beta, \{\vec{x}_i\})) \ p(\{y^*_i\}\  |\  \{(y_i, \vec{x}_i)\}, \beta_m) \ dy_1^*...dy^*_N 
$$

* **Step 4:** Set $\beta_{m+1}$ equal to a value of $\beta$ that maximizes the current approximation $Q(\beta|\beta_m)$ of $\log(p(\{Y^*_i\} \ | \ \beta, \{\vec{x}_i\}))$.

* **Step 5:** Return to step 2 and repeat until some stopping criteria is met.

These formulas may seem difficult at the moment because they are in such a general form. As we specify things for our particular problem things will become more concrete. Now because $Y^*|\vec{X}\sim \mathcal{N}(\ \langle \vec{X}  , \beta\rangle \ ,\ 1)$ is a normal linear regression relationship we have
$$
p(\{Y^*_i\} \ | \ \beta, \{\vec{x}_i\}) = \frac{1}{(2\pi)^{N/2}}e^{\sum_{i=1}^N-\frac{1}{2}\big(Y^*_i - \langle \vec{x}_i  , \beta\rangle\big)^2}
$$

Hence
$$
\log(p(\{Y^*_i\} \ | \ \beta, \{\vec{x}_i\})) = const -\frac{1}{2} \sum_{i=1}^N\big(Y^*_i - \langle \vec{x}_i  , \beta\rangle\big)^2
$$

and so the $Q$-function is
$$
Q(\beta | \beta_m) \ = \ \text{E}_{\{Y^*_i\}\ \ |\ \ \{(Y_i=y_i,\ \vec{X}_i=\vec{x}_i)\},\ \beta_m}  \bigg[ -\frac{1}{2} \sum_{i=1}^N\big(Y^*_i - \langle \vec{x}_i  , \beta\rangle\big)^2 \bigg] + const
$$

Since the data samples are assumed IID we can apply the [representation we derived in the first post in the series](https://edsterjo.netlify.app/post/003_em1/main/) where instead of taking the expectation over all samples, we sum over the expectations of each individual sample:

$$
Q(\beta | \beta_m) \ = \ const - \frac{1}{2}\sum_{i = 1}^N\text{E}_{Y^*_i\ \ |\ \ Y_i=y_i,\ \ \vec{X}_i=\vec{x}_i,\ \ \beta_m}  \bigg[\big(Y^*_i - \langle \vec{x}_i  , \beta\rangle\big)^2 \bigg]
$$

Note that in the EM algorithm we do not actually need to evaluate this function. Instead in step 4 we simply want to find the value of $\beta$ that maximizes it. In addition, as we discussed in the first post, if we wanted to incorporate a prior distribution $p(\beta)$ on $\beta$ for the purpose of regularization we would replace the problem of maximizing $Q(\beta|\beta_m)$ by maximizing $Q(\beta|\beta_m) + \log(p(\beta))$ instead. For the purpose of $L^2$-regularization we could simply take $\beta \sim \mathcal{N}(\vec{0}, \frac{1}{\lambda})$. In that case
$$
Q(\beta|\beta_m) + \log(p(\beta)) = \ const - \frac{1}{2}\sum_{i = 1}^N\text{E}_{Y^*_i\ \ |\ \ Y_i=y_i,\ \ \vec{X}_i=\vec{x}_i,\ \ \beta_m}  \bigg[\big(Y^*_i - \langle \vec{x}_i  , \beta\rangle\big)^2 \bigg] - \frac{\lambda}{2}\langle \beta, \beta\rangle
$$

(the $const$ may now depend on $\lambda$ which we always treat as a constant). We will focus on maximizing this regularized function, knowing that we can simply let $\lambda = 0$ to remove the regularization. At the maximizing point we need the gradient with respect to $\beta$ to equal 0:
$$
\nabla_{\beta} \ \big(Q(\beta|\beta_m) + \log(p(\beta))\big) = 0
$$

(where the $0$ represents the zero vector in $\mathbb{R}^p$). Interchanging gradients first with the summation, then with the expectation (since all random variables have nice distributions) gives
$$
0 = \sum_{i = 1}^N\text{E}_{Y^*_i\ \ |\ \ Y_i=y_i,\ \ \vec{X}_i=\vec{x}_i,\ \ \beta_m}  \bigg[\big(Y^*_i - \langle \vec{x}_i  , \beta\rangle\big)\vec{x}_i \bigg] - \lambda \beta
$$
$$
= \sum_{i = 1}^N\text{E}_{Y^*_i\ \ |\ \ Y_i=y_i,\ \ \vec{X}_i=\vec{x}_i,\ \ \beta_m}  \big[Y^*_i\big]\vec{x}_i - \sum_{i = 1}^N\langle \vec{x}_i  , \beta\rangle\vec{x}_i  - \lambda \beta
$$

Let $X\in\mathbb{R}^{N\times p}$ be an $N\times p$ matrix whose $i$-th row is $\vec{x}_i$, and let $Z\in \mathbb{R}^{N\times 1}$ be a vector with $i$-th component $z_i = \text{E}_{Y^*_i\ \ |\ \ Y_i=y_i,\ \ \vec{X}_i=\vec{x}_i,\ \ \beta_m}  \big[Y^*_i\big]$. Then in matrix notation the above becomes
$$
0 = X^TZ - X^TX\beta - \lambda\beta \\= X^TZ - (X^TX\beta + \lambda I)\beta
$$

This looks very familiar! It looks exactly like the normal equations of OLS if the target variable had been $Z$. The value of $Z = \big(\text{E}_{Y^*_i\ \ |\ \ Y_i=y_i,\ \ \vec{X}_i=\vec{x}_i,\ \ \beta_m}  \big[Y^*_i\big]\big)_{i=1}^N$ is just the value of $Y^*$ we would guess given the values $y_i$ of the censored variable and $\vec{x}_i$ of the covariates. Basically EM is telling us to impute a conditional average for the missing data $Y^*$, fit OLS, and repeat. Solving for $\beta$ gives
$$
\beta_{m+1} = (X^TX + \lambda I)^{-1}X^TZ
$$
This can be implemented once we know the value of $Z$. Since $Z_i = \text{E}_{Y^*_i\ \ |\ \ Y_i=y_i,\ \ \vec{X}_i=\vec{x}_i,\ \ \beta_m}  \big[Y^*_i\big]$ then this is just the mean of a truncated normal distribution:
$$
\text{E}_{Y^*_i\ \ |\ \ Y_i=1,\ \ \vec{X}_i=\vec{x}_i,\ \ \beta_m}  \big[Y^*_i\big] = \langle \vec{x}_i  , \beta_m\rangle + \frac{\phi(\langle \vec{x}_i  , \beta_m\rangle)}{1 - \Phi(-\langle \vec{x}_i  , \beta_m\rangle)}
$$

$$
\text{E}_{Y^*_i\ \ |\ \ Y_i=0,\ \ \vec{X}_i=\vec{x}_i,\ \ \beta_m}  \big[Y^*_i\big] = \langle \vec{x}_i  , \beta_m\rangle - \frac{\phi(\langle \vec{x}_i  , \beta_m\rangle)}{\Phi(-\langle \vec{x}_i  , \beta_m\rangle)}
$$
where $\phi$ and $\Phi$ are the standard normal PDF and CDF respectively. Therefore, we can summarize the EM algorithm for Probit Regression as:

* **Step 1:** Let $m = 0$. Make an initial estimate $\beta_m$ for $\beta$.
* **Step 2:** Impute the censored data according to

$$
z_i =
\begin{cases}
\langle \vec{x}_i  , \beta_m\rangle + \frac{\phi(\langle \vec{x}_i  , \beta_m\rangle)}{1 - \Phi(-\langle \vec{x}_i  , \beta_m\rangle)}, & \text{if} \ \ y_i = 1 \\
\langle \vec{x}_i  , \beta_m\rangle - \frac{\phi(\langle \vec{x}_i  , \beta_m\rangle)}{\Phi(-\langle \vec{x}_i  , \beta_m\rangle)}, & \text{if} \ \ y_i = 0
\end{cases}
$$

* **Step 3:** Solve the regularized OLS problem to update $\beta$:

$$
\beta_{m+1} = (X^TX + \lambda I)^{-1}X^TZ
$$

* **Step 4:** Return to step 2 and repeat until some stopping criteria is met.



## Implementing in C++ using Eigen

The algorithm above is easily implementable in R, Numpy, Matlab, etc., but for fun we'll implement it in C++ using the Eigen linear algebra library. We do this in Rmarkdown using the [RcppEigen](https://cran.r-project.org/web/packages/RcppEigen/index.html) package in R.

```{Rcpp, ref.label=knitr::all_rcpp_labels(), include=FALSE}
```

First we include the necessary header files. Here `RcppEigen.h` includes the Eigen library itself, as well as all the necessary boilerplate code of Rcpp to integrate Eigen (and C++) with R.

```{Rcpp, eval = FALSE}
#include <RcppEigen.h>
#include <algorithm>
#include <limits>

// [[Rcpp::depends(RcppEigen)]]

using namespace Eigen;
```

Next we define two functions that will be needed in computing the vector $Z$, whose computation requires the normal distribution's PDF and CDF. We'll use R's own built-in functions `dnorm` and `pnorm`. These functions are written in C or Fortran (and hence can be called from any language with a C interface) and are very well tested. So instead of rolling our own versions we may as well use R's.

```{Rcpp, eval = FALSE}
// [[Rcpp::export]]
double positive(const double mu) 
{
	double num = R::dnorm(mu, 0, 1, false);
	double den = 1 - R::pnorm(-mu, 0, 1, true, false);
	
	return mu + num/den;
}

// [[Rcpp::export]]
double negative(double mu)
{
	double num = R::dnorm(mu, 0, 1, false);
	double den = R::pnorm(-mu, 0, 1, true, false);
	
	return mu - num/den;
}
```

Next we want a function that actually computes the $Z$ vector, given $X$ and $\beta_m$. We could use some of Eigen's nifty [nullary expressions](https://eigen.tuxfamily.org/dox/TopicCustomizing_NullaryExpr.html), but a simple for-loop with a ternay conditional will do. We also create a wrapper function so that we can test from R:

```{Rcpp, eval = FALSE}
VectorXd impute(const MatrixXd& X,
                const VectorXd& beta,
                const VectorXi& Y)
{
	VectorXd Z = X * beta;
	
	// If Y(i) is non-zero use the function 
	// positive, else use negative
	for(int i = 0; i != Z.size(); ++i)
		Z(i) = (Y(i) == (int)1) ? (positive(Z(i))) : (negative(Z(i)));
	
	return Z;
}

// A wrapper function to test from R
// [[Rcpp::export]]
VectorXd impute_test(const Map<MatrixXd> X,
                     const Map<VectorXd> beta,
                     const Map<VectorXi> Y)
{
	return impute(X, beta, Y);
}
```

Comparing to an R implementation is trivial:
```{r}
X = matrix(rnorm(100*10), ncol = 10)
beta = matrix(rnorm(10), ncol = 1)
Y = matrix(sample(c(0L,1L), size = 100, replace = T), ncol = 1)

# Using the C++ implementation
Z.cpp = impute_test(X, beta, Y)

# Building one in in R
imputeR = function(X, beta, Y)
{
	positiveR = function(Z){Z + dnorm(Z)/(1-pnorm(-Z))}
	negativeR = function(Z){Z - dnorm(Z)/pnorm(-Z)}
	
	Z = X %*% beta
	return(ifelse((Y == 1L), positiveR(Z), negativeR(Z)))
}

Z.r = imputeR(X, beta, Y)

# Checking range of values
print(summary(Z.cpp - Z.r))
```


Not bad. Next we need a function to compute (regularized) least squares. For the solution we use the normal equations. According to the [Eigen tutorial on th matter](https://eigen.tuxfamily.org/dox-devel/group__LeastSquares.html) the normal equations are the fastest but least numerically stable option. For us this is good enough. Again we create a small wrapper to test from R:

```{Rcpp, eval = FALSE}
VectorXd RLS(const VectorXd& Z, 
             const MatrixXd& X, 
             const double lambda = 0.0)
{
	// Creating an identity matrix
	MatrixXd lambda_eye = lambda * MatrixXd::Identity(X.cols(), X.cols());
	
	// Using (regularized) normal equations
	if(lambda >= std::numeric_limits<double>::min())
		return (X.transpose() * X + lambda_eye).ldlt().solve(X.transpose() * Z);
	
	return (X.transpose() * X).ldlt().solve(X.transpose() * Z);
}

// [[Rcpp::export]]
VectorXd RLS_test(const Map<VectorXd> Z, 
                  const Map<MatrixXd> X,
                  double lambda = 0.0)
{
	return RLS(Z, X, lambda);
}
```

Testing in R:
```{r}
set.seed(1234)
X = matrix(rnorm(100*10), ncol = 10)
beta = matrix(rnorm(10), ncol = 1)
Z = X %*% beta + matrix(rnorm(100), ncol = 1)

beta_hat.r = lm.fit(X, Z, method = "qr")$coefficients
beta_hat.cpp = RLS_test(Z, X, 0)

print(data.frame(beta = beta, 
				 beta_hat.r = beta_hat.r, 
				 beta_hat.cpp = beta_hat.cpp))

print(summary(beta_hat.cpp - beta_hat.r))
```
Not bad at all. Note that R's very powerful `lm.fit` function uses QR decomposition to solve the least squares problem. This method is a bit slower in principle than the $LDL^T$ decomposition we used for the normal equations above, but it's also higher quality numerically.

As for a test with a non-trivial regularization constant:

```{r}
lambda = 15
beta_hat_reg.r = solve(t(X) %*% X + lambda*diag(10), t(X) %*% Z)
beta_hat_reg.cpp = RLS_test(Z, X, lambda)

print(data.frame(beta = beta, 
				 beta_hat_reg.r = beta_hat_reg.r, 
				 beta_hat_reg.cpp = beta_hat_reg.cpp))

print(summary(beta_hat_reg.cpp - beta_hat_reg.r))
```
Very good. Again, we see that the regularized least squares estimates are biased away from the true values, and towards the 0 vector.

Now we bring it all together into one algorithm:

```{Rcpp, eval = FALSE}
// [[Rcpp::export]]
VectorXd Probit(const Map<VectorXi> Y, 
                const Map<MatrixXd> X, 
                double lambda = 0.0,
                int num_iter = 100)
{

	// Making sure Lambda is non-negative;
	lambda = std::max(lambda, 0.0);
	
	// Making sure the number of rows of X is the 
	// same as the number of rows of Y
	assert(Y.size() == X.rows());
	
	// Initialize beta to 0 values
	VectorXd beta = VectorXd::Zero(X.cols());
	
	// Iteration
	for(int i = 0; i < num_iter; ++i)
	{
		// Impute the Z vector
		VectorXd Z = impute(X, beta, Y);
		
		// Solve (regularized) Least Squares 
		beta = RLS(Z, X, lambda);
	}
	
	return beta;
}
```

Below we carry out a comparison of the base R implementation of (unregularized) probit regression against our implementation.

```{r, cache=TRUE}
library(MASS)

N = 10000
p = 4

set.seed(1234)
S = matrix(rnorm(p*p), ncol = p)
S = t(S) %*% S

X = MASS::mvrnorm(n = N, mu = rep(0.0, times = p), Sigma = S)

beta = matrix((1:p)/2, ncol = 1)

Z = X %*% beta + matrix(rnorm(N))
Y = as.integer(Z > 0)

system.time(probit.cpp100 <- Probit(Y, X, 0.0, 100))
system.time(probit.cpp10000 <- Probit(Y, X, 0.0, 10000))
system.time(probit.glm <- glm(Y ~ X - 1, family = binomial(link = "probit"))$coefficients)

print(data.frame(beta = beta, 
				 Cpp_100_iter = probit.cpp100,
				 Cpp_10000_iter = probit.cpp10000,
				 R = probit.glm))
```

We see the base R version is much faster, and converges much more quickly. Note this is not due to weak compiler flags on the part of Eigen. A local Makevars file in `Documents/.R` overrides the defaults to use the optimization flags `-O3 -march=native`.

No, instead [`glm.fit`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/glm) uses [iteratively reweighted least squares (IRLS)](https://en.wikipedia.org/wiki/Iteratively_reweighted_least_squares) to fit the model, not EM as can be seen in the [source code](https://github.com/SurajGupta/r-source/blob/master/src/library/stats/R/glm.R). So EM is not a very fast algorithm to fit probit models. In the future we'll implement IRLS in Eigen (or Fortran, or Julia).